**Covid-19 vaccination in the province of Ontario: A geographical and socio-economical analysis**

### 2023-02-14 manuscript/01-background.qmd
updated discussion

Indeed, multiple obstacles that have complicated vaccination efforts against COVID-19 have been identified: inequality in vaccine access, vaccine hesitancy, and differences in vaccination rates across different segments of the population [@gerretsen2021; @nafilyan2021; @malik2020]. In the case of Canada, lower vaccine uptake has been associated with socio-economic factors such as younger age, educational level, presence of children in the household, lack of a regular healthcare provider, ethnic origin, and financial instability [@guay2022;@muhajarine2021;@carter2022;@hussain2022].

>>>>

Indeed, multiple obstacles that have complicated vaccination efforts against COVID-19 have been identified: inequality in vaccine access, vaccine hesitancy, and differences in vaccination rates across different segments of the population [@gerretsen2021; @nafilyan2021; @malik2020]. In the case of Canada, lower vaccine uptake has been associated with socio-economic factors such as younger age, educational level, presence of children in the household, lack of a regular healthcare provider, ethnic origin, and financial instability [@guay2022;@muhajarine2021;@hussain2022].

### 2023-02-14 manuscript/03-results.qmd
updated discussion

data_summary %>%
  modify_header(label = "**Variable**") %>%
  as_kable_extra(booktabs = TRUE) %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down"` argument here.
  kableExtra::kable_styling(latex_options = "repeat_header",font_size = 9)

@@

```{r,model-table}
#| message: false
#| warning: false
#| echo: false
#| label: tbl-model
#| tbl-cap: "Multiple Regression Analysis-Predictors of Vaccination Status"

@@

# create table with the model output
m1 %>%
  tbl_regression(exponentiate = TRUE,
                 label=list(Health_Region ~ "Health Region",
                 Age_group ~ "Age Group")) %>%
    bold_labels()%>%
  as_kable_extra(booktabs = TRUE, longtable=T) %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down" argument here.
  kableExtra::kable_styling(latex_options = "repeat_header")
```

>>>>

data_summary %>%
  modify_header(label = "**Variable**") %>%
  as_kable_extra(booktabs = TRUE,linesep="") %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down"` argument here.
  kableExtra::kable_styling(latex_options = "repeat_header",font_size = 9)

@@

\footnotesize

\renewcommand{\arraystretch}{0.5}
```{r,model-table}
#| message: false
#| warning: false
#| echo: false
#| label: tbl-model
#| tbl-cap: "Multiple Regression Analysis-Predictors of Vaccination Status"

@@

# create table with the model output
m1 %>%
  tbl_regression(exponentiate = TRUE,
                 label=list(Health_Region ~ "Health Region",
                 Age_group ~ "Age Group")) %>%
    bold_labels()%>%
  as_kable_extra(booktabs = TRUE, longtable=T, linesep="") %>%
  # reduce font size to make table fit. 
  # you may also use the `latex_options = "scale_down" argument here.
  kableExtra::kable_styling(latex_options = "repeat_header")
```

### 2023-02-14 manuscript/04-discussion.qmd
updated discussion

Among the different challenges faced by Covid-19 vaccination efforts are the development of new variants due to inadequate public health measures [@li2021] and vaccine inequity  between low and high income countries[@tamey2022]. However, it is also well established that even in the case of high income countries that have had ample access to vaccines since 2020, such as the US, the UK, and Canada, there have been challenges in vaccination efforts due to differences in vaccine uptake among different segments of the population. More specifically, lower vaccine uptake has been associated with socio-economic factors such as race (i.e., identifying as Black, Asian, Indigenous) and household income (typically within lower income brackets) [@willis2021;@skirrow2022;@stoler2021;@khubchandani2021]. Reasons given for this association have included medical mistrust due to systemic medical racism, mistrust in vaccines, and the influence of conspiracy theories [@stoler2021;@willis2021;@bogart2021;@mosby2021;@freeman2020].

@@

- ontario and Health Regions

>>>>

Among the different challenges faced by Covid-19 vaccination efforts are the development of new variants due to inadequate public health measures [@li2021] and inequity in vaccine access between low and high income countries [@tamey2022]. However, it is also well established that even in the case of high income countries that have had ample access to vaccines since 2020, such as the US, the UK, and Canada, there have been challenges in vaccination efforts due to differences in vaccine uptake among different segments of the population. More specifically, lower vaccine uptake has been associated with socio-economic factors such as race (i.e., identifying as Black, Asian, Indigenous) and household income (typically within lower income brackets) [@willis2021;@skirrow2022;@stoler2021;@khubchandani2021]. Reasons given for this association have included medical mistrust due to systemic medical racism, mistrust in vaccines, and the influence of conspiracy theories [@stoler2021;@willis2021;@bogart2021;@mosby2021;@freeman2020].

In addition, vaccine uptake is influenced by geography as shown by different studies that have identified intra-regional differences in vaccine uptake [@pallathadka2022;@huang2022;@mollalo2021]. However, in the case of Canada, most studies that have analyzed spatial differences in vaccination have been focused in country-wide or province-wide estimates [@lavoie2022;@guay2022]. Therefore, we explored spatial and socio-economic determinants of vaccination status in the province of Ontario. This province is of particular interest as it has seen recently major structural health changes with the dissolution of the Local Health Integrated Networks (LHINs) and their incorporation into larger Health Regions [@dong2022]. The idea behind the change aimed at reducing the inequalities in healthcare that were identified under the LHIN approach [@tsasis2012], and thus, examining if there are significant differences in vaccination between the Health Regions could be important to consider by decision-makers in the development of public health policies that aim to address intra-provincial health disparities.

Our results indicate that across the most densely populated Health Regions of Ontario, almost three quarters of the surveyed individuals reported having received the first dose of the Covid-19 vaccine (@tbl-descriptive-stats). This is consistent with overall vaccination rates reported for Canada, which have been relatively higher when compared to other high income countries [@dube2022]. However, we also identified differences in vaccination likelihood based on different socio-economic factors. More specifically, our results show that those that were above 65 years old were only 44% likely to have received a vaccine when compared to those between 16 and 24 years old, that individuals with a household income below CAD 39,999 were around 60% less likely of having received the vaccine when compared to individuals in high income households (with an income above 90,000 CAD), and that individuals who identified as Arab and Middle Easterns, Black, Indigenous, Latin American, South Asian, Southeast Asian, Filipino, West Asian, and minorities not identified elsewhere had significantly lower probability of vaccination  (ranging between 15% and 40%) when compared to White/Caucasian individuals (@tbl-model).

These results are consistent with other studies that have shown an association between lower vaccination rates in racial minorities and lower household income [@guay2022;@muhajarine2021;@carter2022;@hussain2022]. However, it is interesting to note reduced vaccination likelihood in those aged 65 and over, as other studies have reported higher vaccination acceptance in this age group [@macdonald2021;@guay2022].

Although there was no significant difference between vaccination likelihood among the Health Regions, there was a significant lower vaccination likelihood in the Central Health Region in individuals with a household income between CAD 15,000-24,999 (@tbl-model). This result would be consistent in the context of to Ontario Health, The Central Health Region comprises the former Central, Central West, Mississauga Halton, and North Simcoe Muskoka LHINs with 48.3% immigrant population and 47% of the population identifying as visible minority, the highest percentage among all regions [@ontariohealth].

@@

- ontario and Health Regions: find information about poverty and racial groups.
